---
title: "Large Numbers and the Central Limit Theorem"
author: "Martí Cuquet"
output:
  pdf_document:
    fig_caption: yes
---

## Abstract

In this report, we analyze the ToothGrowth data from the R datasets package. We use confidence intervals to compare tooth growth by supplement type and dose.

```{r, echo = FALSE}
library(ggplot2)
library(pander)
```

## Exploratory data analysis

```{r}
data(ToothGrowth)
```

The `ToothGrowth` data set reports on the effect of vitamin C on tooth growth in
guinea pigs. It consists of `r nrow(ToothGrowth)` observations of
`r ncol(ToothGrowth)` variables:

* `len`: The length of odontoblasts (teeth), a numerical variable ranging
  between `r min(ToothGrowth$len)` and `r max(ToothGrowth$len)`.
* `supp`: The supplement type ("VC"" for ascorbic acid or "OJ" for orange
  juice), encoded as a factor.
* `dose`: The dose level in mg, a numerical variable with values 0.5, 1.0 or
  2.0.

To get a bit more insight, we plot in Figure 1 three density plots of the teeth
length depending on the supplement type (OJ or VC) and the dose level (0.5, 1.0
or 2.0 mg).

```{r}
g1 <- ggplot(ToothGrowth, aes(x = len, fill = supp))
g1 <- g1 + geom_density(alpha = 0.6) + facet_wrap(~ dose)
g1 <- g1 + labs(title = "Length of odontoblasts", x = "Length", y = "Density")
g1 <- g1 + theme(text = element_text(size=8))
```

```{r fighistogram, echo = FALSE, fig.cap = "Density plots of the length of odontoblasts. Each panel corresponds to a dose level (0.5, 1.0 or 2.0 mg), red to orange juice supplement and turquoise to ascorbic acid supplement.", fig.width = 7, fig.height = 2.7}
print(g1)
```

At first glance, it seems that orange juice triggers a higher teeth length for
small dosages (0.5 and 1.0 mg), while for a higher dosage (2.0 mg) the results
is basically the same.

We can have a closer look at the mean and variance with respect to dosage and supplement type:

```{r}
ind <- list(ToothGrowth$supp, ToothGrowth$dose)
pantb <- tapply(ToothGrowth$len, ind,
                function(x) sprintf("%.2f(%.2f)", mean(x), sd(x)))
```

This is shown in the following table (standard deviations are shown in
parenthesis):

```{r tabmeans, echo = FALSE, results='asis'}
pandoc.table(pantb)
```

## Tooth growth depending on supplement type and dose

Here we compare tooth growth by supplement type and dosage. The exploratory data
analysis suggests that the growth is higher for orange juice supplement if the
dosage is of either 0.5 or 1 mg, and roughly the same if the dosage is 2 mg. As
we have a small sample, we use Gosset's $t$ confidence intervals to compare the
average growth. We expect it to work well, as the distributions are roughly
symmetric and mound shaped.

We calculate the three 95% confidence intervals of the difference of teeth
length by supplement type corresponding to the three diferent doses.

```{r}
tests <- data.frame(do.call(rbind,
                            lapply(unique(ToothGrowth$dose), function(x) {
                              t.test(len ~ factor(supp), data = ToothGrowth,
                                     paired = TRUE, subset = dose == x)
                              })))
tests <- data.frame(cbind(dose = unique(ToothGrowth$dose),
                          tests[, c("estimate", "conf.int")]))
names(tests) <- c("Dose", "Mean of the differences", "95% confidence interval")
```

The following table shows the result of the tests:

```{r tabtests, echo = FALSE, results='asis'}
pandoc.table(tests)
```

These tests confirmed the observations in the exploratory data analysis. For a
dose of 0.5 or 1 mg, the 95% confidence intervals do not contain 0 and hence
imply that there is a statistically significant difference regarding the
supplement type. More specifically, as the mean of the differences is positive,
in these two cases (dose of 0.5 or 1 mg) orange juice leads to a higher teeth
growth than ascorbic acid. (Note that in the paired test we substracted ascorbic
acid from orange juice.) On the contrary, we can not observe any significant
difference between the supplement types when a dose of 2.0 mg is given.

## Conclusions

In conclusion, there is a statistically significant higher teeth growth for
orange juice as opposed to ascorbic acid supplement, if a dose of 0.5 or 1 mg is
given, while no significant difference exists for a dose of 2 mg. This result
has been obtained assuming roughly symmetric and mound shaped distributions and
using 95% confidence intervals via Gosset's $t$ distribution.